Biden Predicted to Secure a Minimum 10% Lead in Popular Vote Over Trump in the 2024 Election

### 2024-03-16 outputs/paper.qmd
finalize the paper based on peer review and final check

---
title: "Biden Predicted to Secure a Minimum 10% Lead in Popular Votes Over Trump in the 2024 Election"
author: 
  - Boxuan Yi
format: 
  pdf:
    documentclass: article
    geometry: margin = 1in
    abstract: "The dominance of only two political parties in the current US political system adds an extra layer of competitiveness and interest to the presidential election. This paper uses multilevel regression with post-stratification to predict the outcome between the two candidates, Biden and Trump. The model, along with the post-stratification, indicates that Biden is poised to secure a substantial lead of at least 10% in popular votes, suggesting a strong likelihood that the US will continue on a democratic pathway."
date: "11 March 2024"
date-format: "D MMMM YYYY"
number-sections: true
thanks: "Code and data in this analysis is available at: https://github.com/Elaineyi1/2024prediction_USPresidentialElection"
toc: true
bibliography: references.bib
---
```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(knitr)
library(statebins)
library(modelsummary)
library(broom)
library(readr)
library(here)
library(kableExtra)
library(arrow)
```
\newpage
# Introduction
The forthcoming 2024 United States presidential election, scheduled for Tuesday, November 5, 2024 [@270towin270toWin2024], will see US citizens electing to determine the nation’s president and vice president, shaping the pathways of the country for the next four years. The current president Joe Biden, a member of the Democratic Party, is running for re-election, and his predecessor, Donald Trump from the Republican Party, is running for re-election for a nonconsecutive term. Throughout campaigns, political parties articulate their stances and future policies on broader issues such as abortion, immigration, safety, and LGBTQ+ rights. The Democratic and Republican parties typically hold distinct yet steadfast viewpoints, attracting a significant number of supporters among US citizens. However, the post-COVID era has brought about significant changes economically, socially, and politically, making it particularly intriguing to see how these shifts will influence the upcoming 60th quadrennial presidential election.

I will use multilevel regression with post-stratification (MRP) to model and then predict whether Biden or Trump will win the popular votes, which is the total number of votes cast by individual citizens. MRP in this paper involves collecting individual-level survey data from the Polarization Research Lab, analyzing the proportion of support among different demographical groups, and applying the proportion to a larger census dataset from the IPUMS USA, 2022 American Community Survey [@ipumsIPUMS]. One thing that needs to be pointed out is that rather than just depending on the popular votes, the United States elects its president though electoral college. The 51 states are allocated 538 electoral-college votes. The numbers of votes granted to each state, which are different, depend on its representation in the Senate and House of Representatives, and usually every state gives all of its electoral-college votes to whomever wins the popular vote in that state [@economistDoesAmericas]. In other words, it is possible that one candidate wins the popular vote, representing the collective choice of voters, but loses the presidential election because of the distribution of electoral college votes, like the election in 2016 where Hillary Clinton won the popular votes (48%) while losing the election due to the states that supported her have fewer electoral college votes than the ones supporting Trump [@nytimes2016Presidential].

In this paper, I will only predict the popular votes for two candidates, Joe Biden and Donald Trump, utilizing predictors as follows: race, gender, age group, highest level of education, the state of residence. Two datasets will be generated from the survey data —— one being simply cleaned, and the other cleaned and balanced. The latter involves adjusting the survey respondents’ votes in 2020 to align with the actual 2020 results while keeping other characteristics the same. Since the balanced dataset has fewer observations than the original one, this paper focuses on the unbalanced dataset to avoid potential sampling biases. Using the unbalanced dataset, my prediction suggests that Joe Biden is poised to win 62.96% of the popular votes, and 57.61% using the balanced one. Both percentages indicate that Biden is going to secure a minimum 10% lead in the popular votes, which is a pretty large lead in election history.

In this paper, I will delve into the data utilized for analysis, encompassing both survey and census data, in the upcoming section. Following that, I will introduce the model designed to generalize patterns and forecast election outcomes. I will also talk about the predicted results derived from the model. Lastly, I will discuss the results in a broader context and address weaknesses of this paper as well as the future explorations. This paper use the programming language R [@r]. The analysis, the model and the visualizations use the following packages: ipumsr [@ipumsr], dplyr [@rdplyr], readr [@rreadr], tidyr [@rtidyr], arrow [@rarrow], ggplot2 [@rggplot2], knitr [@rknitr], statebins [@rstatebins], modelsummary [@rmodelsummary], broom [@rbroom], here [@rhere], and kableExtra [@r kableExtra].

@@

# Data
## Survey Data and Variables Selection
The survey data I use is from the America’s Political Pulse Survey by the Polarization Research Lab [@polarization]. This open-access online survey contains a series of core questions on politics and special questions designed by political science researchers, posing inquiries to a thousand Americans every week. All the respondents, the population, are paid survey takers from the YouGov survey platform. It collects demographic information, party each respondent support, and core questions like Democrat/Republican thermometer or pride in being American.

The datasets are available weekly. For this paper, I choose 20 weeks of data, spanning from October 6th, 2023 to February 23rd, 2024, forming a dataset with 20000 observations and 69 variables. The variable ‘pid3’ shows the answers to ‘Generally speaking, do you think of yourself as a…?’, with responses being Democratic, Republican, or Independent. For this paper, I will use this variable as supporting the Democratic Party/Biden or Republican Party/Trump for analysis.

In order to predict the 2024 presidential election result using several predictors, I first need to choose variables that can effectively help distinguish between Democrats and Republicans. The survey dataset results show that females tend to support Democracy more than males, with 63.9% female and 58.2% male respondents supporting Democracy. People with higher levels of education tend to support the Democratic Party, especially those who completed post-graduate studies. 55.1% of high school graduates and 60.1% of people with no high school support the Democratic Party, while 72.8% of post-grads support it. More widowed and married people consider themselves as Republicans, while more people who are divorced, never married and separated consider themselves as Democrats. Different ethnicities also show different preferences. Around 74.8% of Asians and 79.7% of African Americans support Biden, compared to 55.7% of white people. Additionally, since the presidential election is voted by state, I would choose race, gender, marital status, highest level of education and state of residence as the variables to explain the binary support for Biden.

The voter turnout rates for different age groups are different. Most of the time, the turnout rate among 60+ years old is the highest, followed by people between 45 to 59, but this is still about 10% higher than the rate among 30-44 years old, and the people from 18 to 29 have the lowest rate of voting for the past 30 years [@ourworldindataElectionVoter]. Hence, I will also use age groups as another factor that affects the popular vote outcome, and the grouping is 18-29, 30-44, 45-59, 60+.

About 22% of individuals identifying with both the Democratic and Republican parties consider voting important, and a significant 70% from each party consider it very important. In this survey, both Democrats and Republicans exhibit similar perceptions of voting importance, indicating that the the survey doesn't inaccurately estimate voting due to differing levels of importance attached to it by the parties. For all the Republican respondents, 55.7% are extremely proud to be Americans, 27.7% are very proud, while only 1.4% are not proud. For all the Democrat respondents, 33.3% are extremely proud, 25.5 are very proud, and 7.3% are not at all proud. Republicans seem to be more polarized as 52.5% of them consider themselves as MAGA Republicans. 22.9% of Democrats think they are very liberal, 36.1% are liberal, while 27.65% of Republicans consider themselves as very conservative, and 44.0% are conservative.

There are also significant differences in religious beliefs. 80% of Republicans think religion is important in their lives (54% very important, 26% somewhat important), while 54.8% of Democrats see it as important (31.1% very important, 23.7% somewhat). Half of the Republican respondents think they are “born-again,” or evangelical Christian, while only 26.5% of Democrats think so. 38.3% of Democrats are ‘Agnostic,’ ‘Atheist,’ or ‘Nothing in particular,’ and 47.6% are ‘Protestant’ or ‘Roman Catholic.’ However, only 17.2% of Republicans are ‘Agnostic,’ ‘Atheist,’ or ‘Nothing in particular,’ and 71.2% are ‘Protestant’ or ‘Roman Catholic.’ Since religion is not collected in the census data, I will not use this factor, but this could be a potential predictor as well.

## Post-Stratification Data
IPUMS USA is a website that contains decennial census from 1790 to 2010, as well as survey data from the American Community Surveys (ACS) and the Puerto Rican Community Surveys from 2000 to the present [@ipumsIPUMS]. The annual ACS public-use samples typically include housing data, demographic data, economic data, and some other individual characteristics. For this analysis, I will use the most recent American Community Survey, the 2022 Sample, as the post-stratification dataset to represent the voter population and adjust the weight of the survey data. This weighted survey used cluster sampling to obtain 1-in-100 national random sample of the population. The population for this sample was the entire US population, categorized into households as subgroups, and the responses of the sampled subgroups were collected.

ACS 2022 dataset consists of 3,373,378 observations, making it a large and relatively representative dataset. The variables I select are gender, age, marital status, race, education and stateicp (ICPSR codes of states) to match with the survey data variables I choose. I then clean and rename the variables to maintain unity with the survey data. [@fig-demographics1] and [@fig-demographics2] display the voters’ demographics in both survey data (pre-strat) and census data (post-strat), with solid lines representing the post-stratification data and dotted lines representing the survey data. The differences in the two lines are due to weight adjustment. Some noticeable adjustment in [@fig-demographics1] include more people aged 30 to 44, more females, fewer married people, fewer asians, and more African American completing the survey. [@fig-demographics2] classifies people based on education and state of residence. There are more people from Florida, New York and Texas in the survey, while states like Wyoming have very few participants, potentially bias.

@@

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-survey_vote
#| fig-cap: Proportion of the Two Parties in the Survey
data_app_clean |>
  group_by(pid3) |>
  summarise(n = n()) |>
  mutate(Proportion = round(n/sum(n)*100)) |>
  ggplot(aes(x = pid3, y = n)) +
  geom_bar(stat = "identity", fill = "grey", color = "black") +
  geom_text(aes(label = paste0(Proportion, "%")), vjust = -0.5, color = "black") +
  labs(x = "Party",
       y = "Number of Respondents") + 
  scale_y_continuous(limits = c(0, 10000)) +
  theme_minimal()
```

@@

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-sample_map
#| fig-cap: Support for Joe Biden and Donald Trump by State in the Sample
survey_map |>
  ggplot(aes(fill = prop, state = statename)) + 
  geom_statebins() + 
  scale_fill_gradient2(low = "#971116", high = "#005D94", mid = "white", midpoint = 0.5) +
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(fill = "Proportion of \nSupporting Biden")
```

## US Election 2020 Dataset and the Balanced Survey Dataset
Among the respondents in the survey, those who voted for Biden in 2020 (44.7%) are approximately 1.5 times as many as those who voted for Trump (30.7%) as shown in [@tbl-survey_votes_2020]. This difference is substantial when compared to the actual popular vote ratio in 2020, which is 51.3%:46.9% [@cookpolitical2020Popular]. I adjust the survey ratio of 2020 votes to match the actual 51.3%:46.9% ratio to create a balanced dataset. To be more specific, I retain all the observations that supported Trump in 2020 but randomly select a certain number of observations that supported Biden in 2020. The number of rows selected is determined by the floor value of 51.3/46.9 times the number of rows that supported Trump.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-survey_votes_2020
#| tbl-cap: Votes in 2020 from the Survey Respondents
data_app_raw |>
  group_by(presvote20post) |>
  summarize(count = n()) |>
  mutate(proportion = count / sum(count) * 100) |> 
  rename('Proportion (%)' = proportion, 
         'Candidates' = presvote20post, 
         'Number of Votes' = count) |> 
  kable()
```

The limitation of this method is that the balanced sample is smaller than the unbalanced sample, causing potential biases and underrepresentation for some groups. The strength is that the balanced dataset contains a more reasonable, rather than a disparate, proportion of supporters. [@fig-balanced_two_compare] (a) shows 3851 respondents consider themselves as Republicans, and 4650 respondents consider themselves as Democrats. [@fig-balanced_two_compare] (b) indicates 3896 people voted for Trump and 4605 voted for Biden in the last election. These two groups of numbers are more balanced.

@@

\newpage
```{r}
#| label: fig-balanced_two_compare
#| fig-cap: Distribution of Supporting Parties Now and Votes in 2020 from the Survey Respondents
#| echo: false
#| layout-ncol: 2
#| warning: false
#| message: false
#| fig-subcap: ["Parties Respondents Support Now", "Respondents' votes in 2020"]
data_app_clean_20 |> mutate(pid3 = factor(pid3, levels = pid3_levels)) |>
  ggplot(aes(x = pid3)) + 
  geom_bar(fill = "grey", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = 2) +
  labs(x = "Party", y = "Number of Counts") +
  theme_minimal()

@@

The data creating [@fig-actual_survey_compare] (a) is a US Election 2020 dataset from Kaggle [@kaggleElection2020], which contains county-level data from the 2020 US Election, including votes, candidates and whether each candidate wins the election. [@fig-actual_survey_compare] shows two maps, one presents the 2020 actual voting results, and one depicts the 2020 voting results from the balanced survey. The distribution of votes from the balanced dataset is more accurate. Alabama (AL) is red in [@fig-actual_survey_compare], Kansas (KS) and Nebraska (NE) are not as blue as in [@fig-sample_map]. I will use the ACS dataset to adjust balanced survey dataset as well. A prediction map using the balanced dataset will be included at the end of the Results section.
```{r}
#| include: false
survey_map_20 <- data_app_clean_20 |>
  group_by(statename, vote_biden) |>
  summarise(n = n()) |>
  group_by(statename) |>
  mutate(prop = n / sum(n)) |>
  filter(vote_biden == "1")
```

```{r}
#| label: fig-actual_survey_compare
#| fig-cap: 2020 Election Votes
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| fig-subcap: ["2020 Actual Voting Results by States", "2020 Voting Results From Balanced Survey Respondents by States"]

# Actual votes in 2020
statevote2020 |> 
  ggplot(aes(fill = support_biden_prop, state = state)) + 
  geom_statebins() + 
  scale_fill_gradient2(low = "#971116", high = "#005D94",
                       mid = "white", midpoint = 0.5) +
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(fill = "Proportion of \nSupporting Biden")

# Survey votes in 2020
survey_map_20 |>
  ggplot(aes(fill = prop, state = statename)) + 
  geom_statebins() + 
  scale_fill_gradient2(low = "#971116", high = "#005D94", mid = "white", midpoint = 0.5) +
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(fill = "Proportion of \nSupporting Biden")
```

@@

# Model
To predict the 2024 election result, I will use Multilevel Regression with Post-Stratification (MRP), a statistical technique that combines the strengths of multilevel modelling and post-stratification. MRP takes a sample, usually a large-scale poll, and uses the poll results to generate a model. This model is then applied to a post-stratification dataset, typically a census or another larger sample [@tellingstorieswithdataTellingStories]. In this paper, the sample is the America’s Political Pulse dataset [@polarization], and the census is the IPUMS ACS dataset [@ipumsIPUMS]. Logistic regression will be used to create the model.

@@

Logistic regression is highly useful when dealing with a binary outcome variable. Instead of using ‘pid3’ variable in the survey, which gives either ‘Democrat’ or ‘Republican’, I create a new variable named ‘vote_biden’. In this new setup, ‘vote_biden’ will be assigned a value of 1 if ‘pid3' is Democrat and 0 if ‘pid3’ is Republican. For instance, if the average "vote_biden" equals 0.5, then half of the population supports Biden.

The logistic regression equation for predicting the binary outcome of support for Biden (where 1 indicates support) can be expressed as follows:
\begin{equation}
\text{logit}(\hat{p}) = \beta_0 + \beta_1 \times \text{gender} + \beta_2 \times \text{race} + \beta_3 \times \text{state} + \beta_4 \times \text{age} + \beta_5 \times \text{marital} + \beta_6 \times \text{education} \label{eq:logit1}
\end{equation}

@@

-  $\text{logit}(\hat{p}) = \log\left(\frac{\hat{p}}{1 - \hat{p}}\right)$
-  $\hat{p}$ is the predicted probability of supporting Biden occurring
-  $\beta_0$ is the intercept term, representing the value if all the predictors are zero
-  $\beta_1$, $\beta_2$, $\beta_3$, $\beta_4$, $\beta_5$, $\beta_6$ are the coefficients corresponding to the predictor variables (gender, race, state, age, marital, education).

@@

```{r}
#| include: false
modelsummary(model)
coefficients <- broom::tidy(model, conf.int = TRUE)
```

>>>>

---
title: "Biden Predicted to Secure a Minimum 10% Lead in Popular Vote Over Trump in the 2024 Election"
author: 
  - Boxuan Yi
format: 
  pdf:
    documentclass: article
    geometry: margin = 1in
    abstract: "The dominance of only two political parties in the current US political system adds an extra layer of competitiveness to the 2024 presidential election. This paper uses multilevel regression with post-stratification to predict the outcome between the two candidates, Biden and Trump. The model, along with the post-stratification, indicates that Biden is poised to secure a substantial lead of at least 10% in popular votes, suggesting a strong likelihood that the US will continue on a democratic pathway."
date: "11 March 2024"
date-format: "D MMMM YYYY"
number-sections: true
thanks: "Code and data in this analysis is available at: https://github.com/Elaineyi1/2024prediction_USPresidentialElection"
toc: true
bibliography: references.bib
---
```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(knitr)
library(statebins)
library(modelsummary)
library(broom)
library(readr)
library(here)
library(kableExtra)
library(arrow)
```
\newpage
# Introduction {#sec-introduction}
The forthcoming 2024 United States presidential election, scheduled for Tuesday, November 5, 2024 [@270towin270toWin2024], will see US citizens electing the nation’s president and vice president, shaping the directions of the country for the next four years. The current president, Joe Biden, a member of the Democratic Party, is running for re-election, and his predecessor, Donald Trump, from the Republican Party, is running for re-election for a nonconsecutive term. During the campaigns, political parties articulate their stances and future policies on broader issues such as abortion, immigration, safety, and LGBTQ+ rights. The Democratic and Republican parties typically hold distinct yet steadfast viewpoints, attracting a significant number of supporters among US citizens. However, the post-COVID era has brought about significant changes economically, socially, and politically, making it particularly intriguing to see how these shifts will influence the upcoming 60th quadrennial presidential election.

I will use multilevel regression with post-stratification (MRP) to model and then predict whether Biden or Trump will win the popular vote, which is the total number of votes cast by individual citizens. MRP in this paper involves collecting individual-level survey data from the Polarization Research Lab [@polarization], analyzing the proportion of support among different demographical groups, and applying the proportion to a larger census dataset from the IPUMS USA, 2022 American Community Survey [@ipumsIPUMS]. One thing needs to be pointed out is that rather than only depending on the popular vote, the United States elects its president though electoral college. The 51 states are allocated 538 electoral-college votes. The numbers of votes granted to each state, which are different, depend on its representation in the Senate and House of Representatives, and usually every state gives all of its electoral-college votes to whomever wins the popular vote in that state [@economistDoesAmericas]. In other words, it is possible that one candidate wins the popular vote, representing the collective choice of voters, but loses the presidential election because of the distribution of electoral college votes, like the election in 2016 where Hillary Clinton won the popular vote (48%) while losing the election because the states that supported her had fewer electoral college votes than those supporting Trump [@nytimes2016Presidential].

In this paper, I will only predict the popular vote for two candidates, Joe Biden and Donald Trump, utilizing predictors as follows: race, gender, age group, highest level of education, the state of residence. Two datasets will be generated from the survey data — one being simply cleaned, and the other cleaned and balanced. The latter involves adjusting the survey respondents’ votes in 2020 to align with the actual 2020 results while keeping other characteristics the same. Since the balanced dataset has fewer observations than the original one, this paper focuses on the result predicted by the unbalanced dataset to avoid potential sampling biases. Based on the unbalanced dataset, my prediction suggests that Joe Biden is poised to win 62.96% of the popular vote, and 57.61% according to the balanced one. Both percentages indicate that Biden is going to secure a minimum 10% lead in the popular vote, which is a pretty large lead in election history.

The data utilized for analysis, including both survey and census data, is shown in @sec-data. Following that, @sec-model introduces the model designed to generalize patterns and forecast election outcomes. I will also talk about the predicted results derived from the model in @sec-results. Lastly, @sec-discussions contains the results in a broader context and address weaknesses of this paper as well as the future explorations. This paper uses the programming language R [@r]. The analysis, the model and the visualizations use the following packages: ipumsr [@ipumsr], dplyr [@rdplyr], readr [@rreadr], tidyr [@rtidyr], arrow [@rarrow], ggplot2 [@rggplot2], knitr [@rknitr], statebins [@rstatebins], modelsummary [@rmodelsummary], broom [@rbroom], here [@rhere], and kableExtra [@rkableExtra].

@@

\newpage
# Data {#sec-data}
## Survey Data and Variables Selection
The survey I use is from the America’s Political Pulse Survey by the Polarization Research Lab [@polarization]. This open-access online survey contains a series of core questions on politics and special questions designed by political science researchers, posing inquiries to one thousand Americans every week. All the respondents, the population, are paid survey takers from the YouGov survey platform. It collects demographic information, party each respondent support, and core questions like Democrat/Republican thermometer or pride in being American.

The datasets are available weekly. For this paper, I choose 20 weeks of data, spanning from October 6th, 2023 to February 23rd, 2024, forming a dataset with 20,000 observations and 69 variables. The variable ‘pid3’ shows the answers to ‘Generally speaking, do you think of yourself as a…?’, with responses being Democratic, Republican, or Independent. For this paper, I will use this variable as supporting the Democratic Party/Biden or Republican Party/Trump.

In order to predict the 2024 presidential election result using several predictors, I first need to choose variables that help effectively differentiate between Democrats and Republicans. The survey dataset results show that more females support the Democratic Party than males, with 63.9% female and 58.2% male respondents supporting Democracy. People with higher levels of education tend to support the Democratic Party, especially those who completed post-graduate studies. 55.1% of high school graduates and 60.1% of people with no high school support the Democratic Party, while 72.8% of post-grads support it. Widowed and married individuals lean Republican, while divorced, never married, and separated people lean Democratic. Ethnicities exhibit varied preferences; around 74.8% of Asians and 79.7% of African Americans support Biden, compared to 55.7% of white individuals. Additionally, since the presidential election is voted by state, I would select race, gender, marital status, highest level of education and state of residence as variables to explain the binary support for Biden. Details about data cleaning can be found in @sec-appendix-cleaning.

The voter turnout rates vary among age groups. Most of the time, the turnout rate among 60+ years old is the highest, followed by people between 45 to 59; but this is still about 10% higher than the rate among 30-44 years old, while individuals from 18 to 29 have the lowest turnout rate for the past 30 years [@ourworldindataElectionVoter]. Hence, I will also select age groups as a factor influencing the popular vote outcome, with groups 18-29, 30-44, 45-59, 60+.

About 22% of individuals identifying with both the Democratic and Republican parties consider voting important, and a significant 70% from each party consider it very important. In this survey, both Democrats and Republicans share similar perceptions of voting importance, implying that the the survey doesn't inaccurately estimate the voting due to different levels of importance perceived by the parties. For all the Republican respondents, 55.7% are extremely proud to be Americans, 27.7% are very proud, while only 1.4% are not proud. For all the Democrat respondents, 33.3% are extremely proud, 25.5 are very proud, and 7.3% are not at all proud. Republicans appear more polarized, with 52.5% identifying as MAGA Republicans. 22.9% of Democrats think they are very liberal, 36.1% are liberal, while 27.65% of Republicans consider themselves as very conservative, and 44.0% are conservative.

There are also significant differences in religious beliefs. 80.0% of Republicans think religion is important in their lives (54.0% very important, 26.0% somewhat important), while 54.8% of Democrats see it as important (31.1% very important, 23.7% somewhat). Half of the Republican respondents think they are “born-again” or evangelical Christian, while only 26.5% of Democrats think so. 38.3% of Democrats are ‘Agnostic’, ‘Atheist’, or ‘Nothing in particular’, and 47.6% are ‘Protestant’ or ‘Roman Catholic’. However, only 17.2% of Republicans are ‘Agnostic’, ‘Atheist’, or ‘Nothing in particular’, and 71.2% are ‘Protestant’ or ‘Roman Catholic’. Since religion is not collected in the census data, I will not use it as a predictor, but it could be considered in future analysis.

## Post-Stratification Data
IPUMS USA is a website that contains decennial census from 1790 to 2010, as well as survey data from the American Community Surveys (ACS) and the Puerto Rican Community Surveys from 2000 to the present [@ipumsIPUMS]. The annual ACS public-use samples typically include housing data, demographic data, economic data, and some other individual characteristics. For this analysis, I will use the most recent American Community Survey, the 2022 Sample, as the post-stratification dataset to represent the voter population. This survey used cluster sampling to obtain 1-in-100 national random sample of the population. The population for this sample was the entire US population, categorized into households as subgroups, and the responses of the sampled subgroups were collected.

ACS 2022 dataset consists of 3,373,378 observations, making it relatively representative. The variables I select are gender, age, marital status, race, education and stateicp (ICPSR codes of states) to match with the survey variables I choose. I then clean and rename the variables to maintain unity with the survey data, details can be found in @sec-appendix-cleaning. [@fig-demographics1] and [@fig-demographics2] display the voters’ demographics in both survey (pre-strat) and census (post-strat), with solid lines representing the post-stratification data and dotted lines representing the survey data. The differences in two lines are due to weight adjustment. Some noticeable adjustment in [@fig-demographics1] include more people aged 30 to 44, more females, fewer married people, fewer asians, and more African American completing the survey. [@fig-demographics2] classifies people based on education and state of residence. There are more people from Florida, New York and Texas in the survey, while states like Wyoming have very few participants, potentially resulting in bias.

@@

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-survey_vote
#| fig-cap: Proportion of the Support for Two Parties in the Survey
data_app_clean |>
  group_by(pid3) |>
  summarise(n = n()) |>
  mutate(Proportion = round(n/sum(n)*100)) |>
  ggplot(aes(x = pid3, y = n)) +
  geom_bar(stat = "identity", fill = "grey", color = "black") +
  geom_text(aes(label = paste0(Proportion, "%")), vjust = -0.5, color = "black") +
  labs(x = "Party",
       y = "Number of Respondents") + 
  scale_y_continuous(limits = c(0, 10000)) +
  theme_minimal()
```

@@

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-sample_map
#| fig-cap: Support for Joe Biden and Donald Trump by State in the Survey
#| fig-subcap: The levels of support for Joe Biden and Donald Trump across different states as reported in the survey
survey_map |>
  ggplot(aes(fill = prop, state = statename)) + 
  geom_statebins() + 
  scale_fill_gradient2(low = "#971116", high = "#005D94", mid = "white", midpoint = 0.5) +
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(fill = "Proportion of Voters \nSupporting Biden")
```

## US Election 2020 Dataset and the Balanced Survey Dataset
Among the respondents in the survey, those who voted for Biden in 2020 (44.7%) are approximately 1.5 times as many as those who voted for Trump (30.7%) as shown in [@tbl-survey_votes_2020]. This difference is substantial when compared to the actual popular vote ratio in 2020, which is 51.3%:46.9% [@cookpolitical2020Popular]. I adjust the survey ratio of 2020 votes to match the actual 51.3%:46.9% ratio to create a balanced dataset. To be more specific, I retain all the observations that supported Trump in 2020 but randomly select a certain number of observations that supported Biden in 2020. This number of rows selected is determined by the floor value of 51.3/46.9 times the number of rows that supported Trump. All figures and tables labeled with 'Balanced' in the caption or title are generated using the balanced dataset. Visualizations without 'Balanced' in the caption use the original, cleaned dataset.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-survey_votes_2020
#| tbl-cap: Votes Cast in 2020 by Respondents of the Survey
data_app_raw |>
  group_by(presvote20post) |>
  summarize(count = n()) |>
  mutate(proportion = count / sum(count) * 100) |> 
  rename('Proportion (%)' = proportion, 
         'Candidates' = presvote20post, 
         'Number of Votes' = count) |> 
  kable()
```

The limitation of this method is that the balanced sample is smaller than the unbalanced sample, causing potential underrepresentation for some groups. The strength is that the balanced dataset contains a more reasonable, rather than a disparate, composition of supporters. [@fig-balanced_two_compare] (a) reveals 3851 respondents consider themselves as Republicans, and 4650 respondents consider themselves as Democrats. [@fig-balanced_two_compare] (b) shows 3896 people voted for Trump and 4605 voted for Biden in the last election. These two groups of numbers are more balanced.

@@

\newpage
```{r}
#| label: fig-balanced_two_compare
#| fig-cap: Parties Supported Now and Votes Cast in 2020 by Survey Respondents (Balanced)
#| echo: false
#| layout-ncol: 2
#| warning: false
#| message: false
#| fig-subcap: ["Parties Survey Respondents Support Now", "Survey Respondents' votes in 2020"]
data_app_clean_20 |> mutate(pid3 = factor(pid3, levels = pid3_levels)) |>
  ggplot(aes(x = pid3)) + 
  geom_bar(fill = "grey", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = 2) +
  labs(x = "Party", y = "Number of Counts") +
  theme_minimal()

@@

The data creating [@fig-actual_survey_compare] (a) is a US Election 2020 dataset from Kaggle [@kaggleElection2020], which contains county-level data from the 2020 US Election, including votes, candidates and whether each candidate wins the election. [@fig-actual_survey_compare] shows two maps, one presents the 2020 actual voting results, and one depicts the 2020 voting results from the balanced survey. The distribution of votes from the balanced dataset is more representative than [@fig-sample_map]. Alabama (AL) is red in [@fig-actual_survey_compare], Kansas (KS) and Nebraska (NE) are not as blue as in [@fig-sample_map]. I will use the post-stratification dataset to adjust the balanced survey as well, and a prediction map using the balanced dataset will be included at the end of @sec-results.
```{r}
#| include: false
survey_map_20 <- data_app_clean_20 |>
  group_by(statename, vote_biden) |>
  summarise(n = n()) |>
  group_by(statename) |>
  mutate(prop = n / sum(n)) |>
  filter(vote_biden == "1")
```

```{r}
#| label: fig-actual_survey_compare
#| fig-cap: 2020 Presidential Election Votes (Balanced)
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| fig-subcap: ["2020 Actual Voting Results by States", "2020 Voting Results by the Balanced Survey Respondents by States"]

# Actual votes in 2020
statevote2020 |> 
  ggplot(aes(fill = support_biden_prop, state = state)) + 
  geom_statebins() + 
  scale_fill_gradient2(low = "#971116", high = "#005D94",
                       mid = "white", midpoint = 0.5) +
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(fill = "Proportion of Voters \nSupporting Biden")

# Survey votes in 2020
survey_map_20 |>
  ggplot(aes(fill = prop, state = statename)) + 
  geom_statebins() + 
  scale_fill_gradient2(low = "#971116", high = "#005D94", mid = "white", midpoint = 0.5) +
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(fill = "Proportion of Voters \nSupporting Biden")
```

@@

# Model {#sec-model}
To predict the 2024 election result, I will use Multilevel Regression with Post-Stratification (MRP), a statistical technique that combines the strengths of multilevel modelling and post-stratification. MRP typically starts with a sample, such as a survey or poll, to generate a model. This model is then applied to a post-stratification dataset, often a census or another larger sample, to produce a more representative result [@tellingstorieswithdataTellingStories]. In this paper, the sample is the America’s Political Pulse dataset [@polarization], and the census is the IPUMS ACS dataset [@ipumsIPUMS]. Logistic regression will be used to create the model.

@@

Logistic regression is highly useful when dealing with a binary outcome variable. I create a new variable assigned a value of 1 if the respondent identifies as a Democrat, and 0 if they identify as Republican. For example, if the average of this new variable equals 0.5, then half of the population supports Biden. The logistic regression equation for predicting the binary outcome of support for Biden (where 1 indicates support) can be expressed as follows:
\begin{equation}
\text{logit}(\hat{p}) = \beta_0 + \beta_1 \times \text{gender} + \beta_2 \times \text{race} + \beta_3 \times \text{state} + \beta_4 \times \text{age} + \beta_5 \times \text{marital} + \beta_6 \times \text{education} \label{eq:logit1}
\end{equation}

@@

-  $\text{logit}(\hat{p}) = \log\left(\frac{\hat{p}}{1 - \hat{p}}\right)$
-  $\hat{p}$ is the predicted probability of supporting Biden occurring
-  $\beta_0$ is the intercept term, representing the value if all the predictors are zero
-  $\beta_1$, $\beta_2$, $\beta_3$, $\beta_4$, $\beta_5$, $\beta_6$ are the coefficients corresponding to the predictor variables, which are gender, race, state, age group, marital status, and education.

@@

```{r}
#| include: false
coefficients <- broom::tidy(model, conf.int = TRUE)
```

I use the binomial distribution to model the probability of the binary outcome in the survey. [@fig-coefficients_model] shows the estimates for the coefficients that will fit into the logistic regression equation, as well as the 95% confidence level for each coefficient. A 95% confidence level means that if I were to repeat the sampling and estimation many times, I would expect the true value of the coefficient to fall within this interval in approximately 95% of repetitions. [@fig-coefficients_model] visualizes the estimates of coefficients as black dots and the confidence intervals as blue lines generated from the model. The length of each blue line reflects the confidence interval for the corresponding coefficient. The zero on the x-axis denotes neutrality. For example, an estimate of -0.1 implies a slight inclination toward voting for Trump. Most data points in [@fig-coefficients_model] are situated to the right of the zero on the x-axis, indicating a higher overall inclination toward supporting Biden. The further to the right both the points and blue lines extend, the greater the likelihood that individuals with these characteristics will vote for Biden. Conversely, points and lines extending to the left suggest a higher probability of supporting Trump. From [@fig-coefficients_model], a person from the District of Columbia is most likely to support Biden, a person from Alaska is most likely to support Trump, and someone who is separated is swing and unpredictable.

After constructing a model, I categorize individuals in both survey and census datasets according to their race, age, gender, marital status, education level and state of residence. I compute the total count and the proportion for each group. Next, utilizing the survey data, I calculate the proportion of voters that tend to support Biden within each demographic group including the confidence intervals, and apply this proportion to the post-stratification dataset for weight adjustment, ensuring a more representative population composition and result.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-coefficients_model
#| fig-cap: Estimates of Coefficients in the Model
coefficients |> 
  ggplot(aes(x = estimate, y = term)) + 
  geom_point() +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), color = "skyblue3", position = position_dodge(width = 0.2)) +
  labs(x = "Estimates and Confidence Intervals", y = "Coefficients of Terms") +
  theme_bw() + 
  theme(axis.text.y = element_text(size = 4.5))
```
